---
title: "DM2-EDA"
output: html_document
date: "`r Sys.Date()`"
---

```{r}
rm(list = ls())
library(readr)
#read dataset
data <- read_csv("dataset/training_v2.csv/training_v2.csv")
dim(data)
```

#Data Preparation/Cleaning
```{r}
#remove the id columns from dataset
data$encounter_id <- NULL
data$hospital_id <- NULL
data$patient_id <- NULL
data$icu_id <- NULL
#drop variables which has null values more than 15%
data <- data[, which(colMeans(is.na(data)) < 0.15)]
dim(data)
```

```{r}
#summary of dataset
summary(data)
#remove because the variable only has a constant value of 0
data$gcs_unable_apache <- NULL
data$readmission_status <- NULL
```

```{r}
library(ggplot2)
ggplot(data, aes(hospital_death)) + 
  geom_histogram()
```

```{r}
#Remove rows with NA's
data_new <- na.omit(data)
print(dim(data_new))
summary(data_new)
```

```{r}
#Combining categories with less than 5% observations into one
data_new$ethnicity[data_new$ethnicity == "Native American" | data_new$ethnicity== "Asian" | data_new$ethnicity == "Hispanic"] <- "Other/Unknown"
#data_new$icu_admit_source[data_new$icu_admit_source == "Other Hospital" | data_new$icu_admit_source == "Other ICU"] <- "other"
#Replacing duplicate factor label with one single label
data_new$apache_2_bodysystem[data_new$apache_2_bodysystem == 'Undefined diagnoses'] <- 'Undefined Diagnoses'

#Factoring
data_new$ethnicity = as.factor(data_new$ethnicity)
data_new$icu_admit_source = as.factor(data_new$icu_admit_source)
data_new$gender = as.factor(data_new$gender)
data_new$icu_stay_type = as.factor(data_new$icu_stay_type)
data_new$apache_2_bodysystem = as.factor(data_new$apache_2_bodysystem)
data_new$icu_type = as.factor(data_new$icu_type)
data_new$apache_3j_bodysystem = as.factor(data_new$apache_3j_bodysystem)
#data_new$hospital_death = as.factor(data_new$hospital_death)
summary(data_new)
```

```{r}
#Save the names of the factor variables as a vector
#Relevel
# vars.cat <- c("ethnicity","gender","icu_type","icu_stay_type", "apache_2_bodysystem","apache_3j_bodysystem")
# 
# for (i in vars.cat){
#     table <- as.data.frame(table(data_new[, i]))
#     max <- which.max(table[, 2])
#     level.name <- as.character(table[max, 1])
#     print(level.name)
#     data_new[, i] <- relevel(data_new[, i], ref = level.name)
# }
```

#Perform oversampling to balance the dataset 
```{r}
#install.packages("ROSE")
library(ROSE)

data_balanced <- ROSE(hospital_death ~ ., data = data_new, hmult.majo = 0, hmult.mino = 1, seed = 1000)$data
summary(data_balanced)
```

```{r}
#Manul binarization
library(caret)
vars.categorical <- c("ethnicity","icu_admit_source","gender","icu_type","icu_stay_type","apache_2_bodysystem","apache_3j_bodysystem")
binarizer <- dummyVars(paste("~", paste(vars.categorical, collapse = "+")),
                        data = data_new, fullRank = TRUE)
binarized_vars <- data.frame(predict(binarizer, newdata = data_balanced))
data_balanced <- cbind(data_balanced, binarized_vars)
data_balanced$ethnicity <- NULL
data_balanced$icu_admit_source <- NULL
data_balanced$gender <- NULL
data_balanced$icu_stay_type <- NULL
data_balanced$icu_type <- NULL
data_balanced$apache_2_bodysystem <- NULL
data_balanced$apache_3j_bodysystem <- NULL
summary(data_balanced)
```

#PCA
```{r}
data_pca <- data_balanced
data_pca$hospital_death <- NULL
data_pca <- data_pca[,1:88]
PCA.scaled <- prcomp(data_pca, center = TRUE, scale = TRUE)

#proportion of variance explained by each principal component
variance_prop <- PCA.scaled$sdev^2 / sum(PCA.scaled$sdev^2)
cum_variance_prop <- cumsum(variance_prop)
num_pcs <- sum(cum_variance_prop <= 0.5)
PCA.selected <- PCA.scaled$x[,1:num_pcs]
PCA.selected
```

```{r}
data_balanced[2:89] <- NULL
data_balanced <- cbind(data_balanced, PCA.selected)
summary(data_balanced)
```

```{r}
library(caret)
set.seed(1000)
partition <- createDataPartition(data_balanced$hospital_death, p = 0.75, list = FALSE)
data.train <- data_balanced[partition, ]
data.test <- data_balanced[-partition, ]
mean(data.train$hospital_death)
mean(data.test$hospital_death)
```
#Regularization
```{r}
#library(glmnet)

# X.train <- model.matrix(hospital_death ~ ., data = data.train)
# 
# set.seed(1111)
# m <- cv.glmnet(x = X.train,
#               y = data.train$hospital_death,
#               family = "binomial",
#               nfold = 10,
#               #lambda = NULL,
#               lambda = 1:20,
#               alpha = 0.5)
# plot(m)
```

```{r}
library("dplyr")
#generate correlation plot
#data.numeric <- data[,unlist(lapply(data, is.numeric))]
# cor.matrix <- cor(data.numeric[1:10], use = "complete.obs")
# data <- round(cor.matrix, digits=4)
# print(data)
```

#Fitting Models
#GLM Train
```{r}
model.logit.full <- glm(hospital_death ~ ., 
                         data = data.train,
                         family = binomial(link = "logit"))
AIC(model.logit.full)
summary(model.logit.full)
```
#GLM Predict
```{r}
cutoff <- 0.5

# Generate predicted probabilities on the training set
predict.logit.train <- predict(model.logit.full, type = "response")
# Turn predicted probabilities into predicted classes
class.train <- 1*(predict.logit.train > cutoff) # OR ifelse(pred.train > cutoff, 1, 0)
confusionMatrix(factor(class.train), factor(data.train$hospital_death), positive = "1")

predict.logit.test <- predict(model.logit.full, newdata = data.test, type = "response")
class.logit <- 1*(predict.logit.test > cutoff)
confusionMatrix(factor(class.logit), factor(data.test$hospital_death))
```
The training accuracy is slightly higher than the testing accuracy.
This is an indication of overfitting but not significant. 
The sensitivity is higher meaning that the model has a higher capability of
classifying patients who died.This is our parimary objective of this model
# ROC AUC
```{r}
library(pROC)
roc.logit.train <- roc(data.train$hospital_death, predict.logit.train)
par(pty = "s")  # This improves the appearance of the roc plot
plot(roc.logit.train)
auc(roc.logit.train)

roc.logit.test <- roc(data.test$hospital_death, predict.logit.test)
plot(roc.logit.test)
auc(roc.logit.test)
```
# Stepwise AIC
```{r}
library(MASS)
model.logit.null <- glm(hospital_death ~ 1,
                  data = data.train,
                  family = binomial(link = "logit"))
# model.logit.reduced <- stepAIC(model.logit.null,
#                          direction = "backward",
#                          k = log(nrow(data.train)),
#                          scope = list(upper = model.logit.full, lower = model.logit.null))
model.logit.reduced <- stepAIC(model.logit.full)
summary(model.logit.reduced)
```

# GLM Reduced 
```{r}
predict.logit.reduced <- predict(model.logit.reduced, newdata = data.test, type = "response")
roc.logit.reduced <- roc(data.test$hospital_death, predict.logit.reduced)
auc(roc.logit.reduced)
```