---
title: "DM2-PCA-biplot-investigation"
output: html_document
date: "`r Sys.Date()`"
---

```{r}
rm(list = ls())
library(readr)
#read dataset
data <- read_csv("training_v2.csv")
dim(data)
```

#Data Preparation/Cleaning
```{r}
#remove the id columns from dataset
data$encounter_id <- NULL
data$hospital_id <- NULL
data$patient_id <- NULL
data$icu_id <- NULL
#drop variables which has null values more than 15%
data <- data[, which(colMeans(is.na(data)) < 0.15)]
dim(data)
```

```{r}
#summary of dataset
summary(data)
#remove because the variable only has a constant value of 0
data$gcs_unable_apache <- NULL
data$readmission_status <- NULL
data$apache_4a_hospital_death_prob <- NULL
data$apache_4a_icu_death_prob <- NULL
```

```{r}
library(ggplot2)
ggplot(data, aes(hospital_death)) + 
  geom_histogram()
```

```{r}
#Remove rows with NA's
data_new <- na.omit(data)
print(dim(data_new))
summary(data_new)
```

```{r}
#Combining categories with less than 5% observations into one
data_new$ethnicity[data_new$ethnicity == "Native American" | data_new$ethnicity== "Asian" | data_new$ethnicity == "Hispanic"] <- "Other/Unknown"
#data_new$icu_admit_source[data_new$icu_admit_source == "Other Hospital" | data_new$icu_admit_source == "Other ICU"] <- "other"
#Replacing duplicate factor label with one single label
data_new$apache_2_bodysystem[data_new$apache_2_bodysystem == 'Undefined diagnoses'] <- 'Undefined Diagnoses'

#Factoring
data_new$ethnicity <- as.factor(data_new$ethnicity)
data_new$icu_admit_source <- as.factor(data_new$icu_admit_source)
data_new$gender <- as.factor(data_new$gender)
data_new$icu_stay_type <- as.factor(data_new$icu_stay_type)
data_new$apache_2_bodysystem <- as.factor(data_new$apache_2_bodysystem)
data_new$icu_type <- as.factor(data_new$icu_type)
data_new$apache_3j_bodysystem <- as.factor(data_new$apache_3j_bodysystem)
data_new$leukemia <- as.factor(data_new$leukemia)
data_new$lyphoma <- as.factor(data_new$lymphoma)
data_new$cirrhosis <- as.factor(data_new$cirrhosis)
data_new$diabetes_mellitus <- as.factor(data_new$diabetes_mellitus)
data_new$hepatic_failure <- as.factor(data_new$hepatic_failure)
data_new$immunosuppression <- as.factor(data_new$immunosuppression)
data_new$aids <- as.factor(data_new$aids)
data_new$arf_apache <- as.factor(data_new$arf_apache)
#data_new$hospital_death = as.factor(data_new$hospital_death)
summary(data_new)
```

#Perform oversampling to balance the dataset 
```{r}
#install.packages("ROSE")
library(ROSE)

#data_balanced <- ROSE(hospital_death ~ ., data = data_new, seed = 1000)$data
data_balanced <- ovun.sample(hospital_death ~ .,
                             data = data_new,
                             N = nrow(data_new),
                             seed = 1000,
                             method = "both")$data
                             
summary(data_balanced)
```

# PCA Biplot, intial PCA and loadings -MO
```{r}

library(dplyr)
data_pca <- data_balanced
data_pca <- select_if(data_pca, is.numeric)


data_pca$hospital_death <- NULL
PCA <- prcomp(data_pca, center=TRUE, scale. = TRUE)

# Sort results based on PC1 (first column) in ascending order
PCA$rotation <- PCA$rotation[order(PCA$rotation[, 1]), ]

# Display results for the first two principal components
PCA$rotation[, 1:2]

# Sort results based on PC2 (second column) in ascending order
PCA$rotation <- PCA$rotation[order(PCA$rotation[, 2]), ]

# Display results for the first two principal components
PCA$rotation[, 1:2]
```

# prep for view -MO
```{R}
# set above into a table
PCA_rotation <- read.table(text = "PC1          PC2
h1_mbp_noninvasive_min      -2.272161e-01 -0.034004790
h1_mbp_min                  -2.271627e-01 -0.033892130
h1_sysbp_noninvasive_min    -2.155549e-01 -0.047903059
h1_sysbp_min                -2.151376e-01 -0.048902345
h1_diasbp_noninvasive_min   -2.110806e-01 -0.029536258
h1_diasbp_min               -2.108560e-01 -0.028399677
h1_mbp_noninvasive_max      -2.103472e-01  0.137194174
h1_mbp_max                  -2.095512e-01  0.138568370
h1_sysbp_noninvasive_max    -2.019668e-01  0.111721757
h1_sysbp_max                -2.015528e-01  0.110573129
d1_mbp_min                  -1.892799e-01 -0.159669148
d1_mbp_noninvasive_min      -1.892575e-01 -0.159907427
d1_sysbp_noninvasive_min    -1.837676e-01 -0.167355916
d1_sysbp_min                -1.837321e-01 -0.167549618
h1_diasbp_noninvasive_max   -1.815893e-01  0.163498322
h1_diasbp_max               -1.807889e-01  0.163574902
d1_diasbp_min               -1.783949e-01 -0.149379405
d1_diasbp_noninvasive_min   -1.782589e-01 -0.149842415
d1_sysbp_noninvasive_max    -1.634758e-01  0.170161642
d1_sysbp_max                -1.634615e-01  0.169951185
d1_mbp_noninvasive_max      -1.607105e-01  0.210241745
d1_mbp_max                  -1.604116e-01  0.210265588
map_apache                  -1.369629e-01  0.088759061
d1_diasbp_max               -1.306288e-01  0.224218024
d1_diasbp_noninvasive_max   -1.306156e-01  0.224519777
d1_hemaglobin_min           -1.095219e-01 -0.006444698
d1_hematocrit_min           -1.044549e-01  0.001014823
d1_calcium_min              -1.023703e-01 -0.096840892
d1_hemaglobin_max           -9.323971e-02  0.031046936
d1_hematocrit_max           -8.989695e-02  0.038862784
d1_calcium_max              -8.825140e-02 -0.043393197
d1_spo2_min                 -6.170293e-02 -0.143896903
d1_platelets_min            -4.290331e-02 -0.022893040
h1_spo2_min                 -4.279932e-02 -0.105521619
gcs_verbal_apache           -3.702390e-02 -0.181892947
gcs_motor_apache            -3.175163e-02 -0.180468388
gcs_eyes_apache             -3.158434e-02 -0.184349443
d1_platelets_max            -3.084889e-02  0.004271328
d1_temp_min                 -2.947639e-02 -0.112953428
apache_2_diagnosis          -2.896623e-02 -0.087518320
temp_apache                 -2.520019e-02 -0.083909929
height                      -2.337329e-02 -0.002626387
weight                      -2.210372e-02  0.022228865
d1_glucose_min              -1.539222e-02  0.027741190
h1_spo2_max                 -1.388278e-02 -0.027158633
d1_sodium_min               -1.272149e-02  0.023299702
bmi                         -1.156606e-02  0.024176761
d1_resprate_min             -8.445898e-03 -0.029807510
elective_surgery             2.849049e-05 -0.095589759
apache_post_operative        9.056545e-04 -0.091794862
lymphoma                     5.290620e-03  0.003885919
resprate_apache              5.653509e-03  0.075896991
glucose_apache               5.948471e-03  0.099861636
d1_sodium_max                6.174068e-03  0.081764990
apache_3j_diagnosis          8.348943e-03 -0.092685411
d1_temp_max                  9.916260e-03  0.038854893
d1_spo2_max                  1.013326e-02  0.015024598
d1_heartrate_min             1.021821e-02 -0.014025704
solid_tumor_with_metastasis  1.150804e-02  0.003481843
h1_resprate_min              1.316370e-02  0.059542752
d1_glucose_max               1.650624e-02  0.112443541
h1_resprate_max              2.028956e-02  0.135703393
d1_resprate_max              2.425838e-02  0.129269882
pre_icu_los_days             2.515240e-02  0.003691603
h1_heartrate_min             2.703050e-02  0.099128218
d1_potassium_min             3.214345e-02 -0.006058285
h1_heartrate_max             3.404007e-02  0.158953387
heart_rate_apache            3.607979e-02  0.135474021
age                          3.678042e-02  0.008139273
d1_wbc_min                   3.692917e-02  0.077815276
intubated_apache             4.158519e-02  0.139661712
ventilated_apache            4.884172e-02  0.168623998
d1_wbc_max                   4.957457e-02  0.100547644
d1_heartrate_max             4.992530e-02  0.189244997
d1_potassium_max             5.799947e-02  0.084373328
d1_creatinine_min            5.946960e-02  0.076288601
d1_creatinine_max            6.329757e-02  0.091840521
d1_bun_min                   8.679778e-02  0.077723952
d1_bun_max                   8.953100e-02  0.091433096", header = TRUE)

```

# initial plot -MO
```{R}
# Load the required library
library(ggplot2)

# Create a biplot of the first two principal components
biplot(PCA, scale = 0)

```
# PCA-biplot display -MO
# shown without observations, and expanded, for clarity -MO
```{r}
# Load the required library
library(ggplot2)

# Subset and label only top variables (you can adjust 'n' to show more or fewer variables)
n <- 900
top_vars <- head(PCA_rotation[order(-abs(PCA_rotation$PC1)), ], n)

# Create a biplot using ggplot2
ggplot(data = PCA_rotation, aes(x = PC1, y = PC2)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  geom_segment(aes(xend = 0, yend = 0, x = PC1, y = PC2),
               arrow = arrow(length = unit(0.3, "cm")), color = "blue") +
  geom_text(data = top_vars, aes(label = rownames(top_vars)), vjust = 1, color = "red", size = 3) +
  xlim(min(PCA_rotation$PC1) * 1.1, max(PCA_rotation$PC1) * 1.1) +
  ylim(min(PCA_rotation$PC2) * 1.1, max(PCA_rotation$PC2) * 1.1) +
  xlab("PC1") +
  ylab("PC2") +
  theme_minimal()

```

# Show the observations on the Bi-plot -MO
```{R}
# Load the required libraries
library(ggbiplot)
library(devtools)

# Create a biplot of the first two principal components
ggbiplot(PCA, scale = 0, obs.scale = 0, varname.size = 3, varname.adjust = 0.7)

```

# set up for pca bi-plot with data overlay -MO
```{R}
# Load required libraries
library(stats)
library(dplyr)

# this is for the "unbalanced" and original data that has not been oversampled
data_pca_un <- data_new
data_pca_un_n <- select_if(data_pca_un, is.numeric)

# take out the hospital_death for PCA
data_pca_un_n$hospital_death <- NULL
PCA_orig <- prcomp(data_pca_un_n, center=TRUE, scale. = TRUE)

# set x an y
X_pca_un <- as.data.frame(PCA_orig$x)  # Convert the matrix to a data frame
y_un <- data_pca_un$hospital_death


# this is for the "balanced" data 

# Extract the PCA-transformed features
X_pca <- as.data.frame(PCA$x)  # Convert the matrix to a data frame

# Display the first few rows of the PCA-transformed features
head(X_pca)

# Replace "target_variable_column_name" with the actual name of your target variable column
y <- data_balanced$hospital_death

# Display the first few elements of the target variable
head(y)

```

# PC biplot with overlay of deaths and counts (after balance and oversampling) -MO
```{R}
# Load required libraries
library(stats)
library(ggplot2)

# 'X_pca' contains the PCA-transformed features, 'y' contains the target variable
# 'X_pca' should have dimensions (n_samples, n_components), where n_components is typically 2 for a biplot

# Combine PCA-transformed features and target variable into a data.frame
pca_data_df2 <- data.frame(X_pca, hospital_death = factor(y))

# Calculate the counts for each class
counts <- table(pca_data_df2$hospital_death)

# Create a bi-plot using ggplot2
biplot <- ggplot(pca_data_df2, aes(x = PC1, y = PC2, color = hospital_death)) +
  geom_point(alpha = 0.5) +
  labs(x = "Principal Component 1", y = "Principal Component 2") +
  scale_color_manual(
    values = c("blue", "red"),
    labels = c(
      paste("hospital_death = 0 (", counts[1], ")", sep = ""),
      paste("hospital_death = 1 (", counts[2], ")", sep = "")
    )
  ) +
  theme_minimal() +
  ggtitle("PCA Bi-Plot with hospital_death Overlay - after Balancing")

# Show the bi-plot
print(biplot)

```
# PC biplot with overlay of deaths and counts (original, unbalanced data) -MO
```{R}
# Load required libraries
library(stats)
library(ggplot2)

# 'X_pca_un' contains the PCA-transformed features, 'y_un' contains the target variable
# 'X_pca_un' should have dimensions (n_samples, n_components), where n_components is typically 2 for a biplot

# Combine PCA-transformed features and target variable into a data.frame
pca_data_df3 <- data.frame(X_pca_un, hospital_death = factor(y_un))

# Calculate the counts for each class
counts <- table(pca_data_df3$hospital_death)

# Create a bi-plot using ggplot2
biplot <- ggplot(pca_data_df2, aes(x = PC1, y = PC2, color = hospital_death)) +
  geom_point(alpha = 0.5) +
  labs(x = "Principal Component 1", y = "Principal Component 2") +
  scale_color_manual(
    values = c("blue", "red"),
    labels = c(
      paste("hospital_death = 0 (", counts[1], ")", sep = ""),
      paste("hospital_death = 1 (", counts[2], ")", sep = "")
    )
  ) +
  theme_minimal() +
  ggtitle("PCA Bi-Plot with hospital_death Overlay - unbalanced/original")

# Show the bi-plot
print(biplot)
```